Beating Simple Backtracking Search in Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1].}

### 2018-11-28 paper/paper.tex
minor fixes to paper

Although the problem is \NP-complete, by combining design techniques from artificial intelligence
with careful algorithm engineering, modern subgraph isomorphism solvers can often produce exact
solutions quickly even on graphs with thousands of vertices. The current single strongest subgraph
isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to direct a
constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. To offset mistakes
made by this heuristic, this paper proposes a new perspective on value-ordering: rather than
defining a search order, we use degree to direct what \emph{proportion} of the search effort should
be spent in each subproblem. By combining rapid restarts and nogood recording, and introducing a
small amount of randomness into the value-ordering heuristic, we make a state-of-the-art subgraph
algorithm perform two orders of magnitude better on a large number of satisfiable instances, whilst
performing worse only rarely on satisfiable instances, and never on unsatisfiable instances.  This
strategy is also effective in an optimisation setting, producing benefits in two maximum common
induced subgraph algorithms.

This new form of search can also be parallelised, with a \emph{much} simpler implementation than
conventional work-stealing. By running many threads with different random seeds but the same restart
schedule, and sharing nogoods only following restarts, we can achieve speedups of thirty one from a
thirty six core machine, or ninety five by using ten such machines with message passing.

@@

This paper looks at improving the Glasgow Subgraph
Solver\footnote{\url{https://github.com/ciaranm/glasgow-subgraph-solver/}}, which can solve both the
non-induced and the induced subgraph isomorphism problems.  The solver is very closely based upon
the $k{\downarrow}$ algorithm of \citet{Hoffmann et al.\ }{DBLP:conf/aaai/HoffmannMR17} with $k =
0$, and we refer the reader to that paper for full technical details; that algorithm, in turn, is a
simplification and re-engineering of an older Glasgow algorithm
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}.  Essentially, the solver is essentially
a dedicated forward-checking constraint programming implementation specifically for subgraph
problems. It works with a model having a variable per pattern graph vertex, with domains ranging
over the target graph vertices, and performs a backtracking search to map each pattern vertex to a
target vertex whilst propagating adjacency and injectivity constraints (together with further
implied constraints based upon degrees and paths).  However, it uses specialised bit-parallel data
structures and algorithms, and propagates constraints in a fixed order rather than using a queue.

@@

\begin{description}
    \item[BVG(r), M4D(r), and Rand] are families of randomly generated graphs using different models (bounded
        degree, regular mesh, and uniform), where each pattern is a permuted random connected subgraph
        of the target (and so each instance is satisfiable) \cite{DBLP:journals/pami/CordellaFSV04}.
        These benchmark instances are widely used, but have unusual properties and so broad
        conclusions should not be drawn based solely upon behaviour on these instances
        \cite{DBLP:journals/jair/McCreeshPST18}.
    \item[SF] contains randomly generated scale-free graphs using a similar method
        \cite{DBLP:journals/constraints/ZampelliDS10}.
   \item[LV] consists of various kinds of graph gathered by Larrosa and Valiente
       \cite{DBLP:journals/mscs/LarrosaV02} from the Stanford Graph Database. We include both the
        50 small graphs, and the 50 larger graphs.
    \item[Phase] contains hard crafted instances that lie near the satisfiable / unsatisfiable phase
        transition \cite{DBLP:journals/jair/McCreeshPST18}.
    \item[PR] contains graphs generated from segmented images, corresponding to a computer vision
        problem \cite{DBLP:journals/pr/SolnonDHJ15}.
    \item[Images and Meshes] contain graphs representing 2D segmented images and 3D object models,
        again representing a computer vision problem \cite{DBLP:journals/cviu/DamiandSHJS11}.
\end{description}

@@

To avoid exploring portions of the search
space that we have already visited, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees---such a constraint is called a nogood. We generate
simple decision nogoods. That is, upon backtracking due to a decision to restart, we post a nogood
of the form $(v \mapsto v') \wedge (w \mapsto w') \wedge (x \mapsto x') \Rightarrow \bot$ for every
branch to the left of the current (incomplete) branch in the search tree, and when we first make a
decision to restart before backtracking, we post a similar nogood eliminating the entire subtree
explored. We use the two watched literals technique \cite{DBLP:conf/dac/MoskewiczMZZM01} to
propagate stored nogoods.
%(\siplineref{2wl}).
This has two benefits: the propagation complexity does not particularly depend upon the number of
stored nogoods, and it does not require any work upon backtracking.  Other more sophisticated nogood
generation and propagation schemes exist \cite{DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17},
but these are not helpful in this setting (our solver does not maintain arc consistency or use a
propagation queue).

@@

Solution-biased search allows for a much simpler parallel implementation, which we now evaluate.  We
create a number of threads, and give each thread its own random seed, but otherwise each thread
performs the same sequential search. Then, whenever a thread restarts, it uses a barrier to wait for
every other thread to also restart. Each thread's new nogoods from that restart can then be
gathered, and then each thread resumes, with a larger set of nogoods than would have been generated
in a sequential run. Finally, whenever any single thread terminates, either due to having found a
solution or proved unsatisfiability, then every other thread may immediately terminate.

@@

\begin{itemize}
    \item Each thread will be run with the same restart schedule, and so will spend approximately
        the same amount of time between restarts. Because the only synchronisation between threads
        is at restarts, we expect threads to be kept busy doing useful work.  (This is in contrast
        to an alternative method for paralellising restarting search \cite{DBLP:conf/aaai/CireKS14},
        which packs together successive sequence values to produce a balanced workload.)
    \item Sequentially, on non-trivial instances the algorithm will restart very many times (many
        tens of thousands, for instances that reach the thousand second timeout).
    \item Because the search trees we explore are exponentially large, the randomness in the
        value-ordering heuristic is sufficient to ensure that most of the time, threads are
        exploring different parts of the search tree.
    \item The gathering of nogoods to describe the work done so far provides an alternative to
        requiring either a specific mechanism to allocate work, or expensive synchronisation between threads.
    \item If sometimes threads do happen to explore part of
        the same subproblems, this is not a problem: if the instance is satisfiable, either thread
        might find a solution first, and if the instance is unsatisfiable, we merely introduce some
        redundancy into the proof.\footnote{For solving a counting or enumeration problem,
        matters become slightly more complicated, but not devastatingly so.} The combination of
        rapid restarts and nogood recording is enough to ensure that this is only a small overhead.
\end{itemize}

@@

It is important to emphasise that there is no expectation of a linear speedup from this approach,
particularly for satisfiable instances. Due to the random nature of the value-ordering heuristic,
one of our extra threads may ``get lucky'' and find a solution much quicker than we would
sequentially, leading to a superlinear speedup. Conversely, our extra threads may contribute nothing to finding a
solution---or worse, due to new nogoods altering the choice made by the random branching heuristic,
we could even get an absolute slowdown. We can see both of these effects in the bottom left plot of
\cref{figure:parallel}: although we achieve roughly a linear speedup on unsatisfiable instances,
satisfiable instances show much greater variability. However, this approach does at least mirror our intuition of
allocating search effort in proportion to where the value-ordering heuristic believes it will be
most fruitful, and so we should not be too surprised that we see roughly a linear speedup on average
on harder instances.

@@

The authors would like to thank Ozgur Akgun, Chris Jefferson, Christophe Lecoutre, Christine
Solnon, and Craig Reilly for their comments.

>>>>

Although the problem is \NP-complete, by combining design techniques from artificial intelligence
with careful algorithm engineering, modern subgraph isomorphism solvers can often produce exact
solutions quickly even on graphs with thousands of vertices. The current single strongest subgraph
isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to direct a
constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. To offset mistakes
made by this heuristic, this paper proposes a new perspective on value-ordering: rather than
defining a search order, we use degree to direct what \emph{proportion} of the search effort should
be spent in each subproblem. By combining rapid restarts and nogood recording, and introducing a
small amount of randomness into the value-ordering heuristic, we make a state-of-the-art subgraph
algorithm perform two orders of magnitude better on a large number of satisfiable instances, whilst
performing worse rarely on satisfiable instances, and never on unsatisfiable instances.  This
strategy is also effective in an optimisation setting, producing benefits in two maximum common
induced subgraph algorithms.

This new form of search can also be parallelised, with a much simpler implementation than
conventional work-stealing. By running many threads with different random seeds but the same restart
schedule, and sharing nogoods only following restarts, we can achieve speedups of thirty one from a
thirty six core machine, or ninety five by using ten such machines with message passing.

@@

This paper looks at improving the Glasgow Subgraph
Solver\footnote{\url{https://github.com/ciaranm/glasgow-subgraph-solver/}}, which can solve both the
non-induced and the induced subgraph isomorphism problems.  The solver is very closely based upon
the $k{\downarrow}$ algorithm of \citet{Hoffmann et al.\ }{DBLP:conf/aaai/HoffmannMR17} with $k =
0$, and we refer the reader to that paper for full technical details; that algorithm, in turn, is a
simplification and re-engineering of an older Glasgow algorithm
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}.  Essentially, the solver is
a dedicated forward-checking constraint programming implementation specifically for subgraph
problems. It works with a model having a variable per pattern graph vertex, with domains ranging
over the target graph vertices, and performs a backtracking search to map each pattern vertex to a
target vertex whilst propagating adjacency and injectivity constraints (together with further
implied constraints based upon degrees and paths).  However, it uses specialised bit-parallel data
structures and algorithms, and propagates constraints in a fixed order rather than using a queue.

@@

\begin{description}
    \item[BVG(r), M4D(r), and Rand] are families of randomly generated graphs using different models (bounded
        degree, regular mesh, and uniform), where each pattern is a permuted random connected subgraph
        of the target (and so each instance is satisfiable) \cite{DBLP:journals/pami/CordellaFSV04}.
        These benchmark instances are widely used, but have unusual properties and so broad
        conclusions should not be drawn based solely upon behaviour of these instances
        \cite{DBLP:journals/jair/McCreeshPST18}.
    \item[SF] contains randomly generated scale-free graphs using a similar method
        \cite{DBLP:journals/constraints/ZampelliDS10}.
   \item[LV] consists of various kinds of graph gathered by Larrosa and Valiente
       \cite{DBLP:journals/mscs/LarrosaV02} from the Stanford Graph Database. We include both the
        50 small graphs, and the 50 large graphs.
    \item[Phase] contains hand crafted instances that lie near the satisfiable / unsatisfiable phase
        transition \cite{DBLP:journals/jair/McCreeshPST18}.
    \item[PR] contains graphs generated from segmented images, corresponding to a computer vision
        problem \cite{DBLP:journals/pr/SolnonDHJ15}.
    \item[Images and Meshes] contain graphs representing 2D segmented images and 3D object models,
        again representing a computer vision problem \cite{DBLP:journals/cviu/DamiandSHJS11}.
\end{description}

@@

To avoid exploring portions of the search
space that we have already visited, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees---such a constraint is called a \emph{nogood}. We generate
simple decision nogoods. That is, upon backtracking due to a decision to restart, we post a nogood
of the form $(v \mapsto v') \wedge (w \mapsto w') \wedge (x \mapsto x') \Rightarrow \bot$ for every
branch to the left of the current (incomplete) branch in the search tree, and when we first make a
decision to restart before backtracking, we post a similar nogood eliminating the entire subtree
explored. We use the two watched literals technique \cite{DBLP:conf/dac/MoskewiczMZZM01} to
propagate stored nogoods.
%(\siplineref{2wl}).
This has two benefits: the propagation complexity does not particularly depend upon the number of
stored nogoods, and it does not require any work upon backtracking.  Other more sophisticated nogood
generation and propagation schemes exist \cite{DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17},
but these are not helpful in this setting (our solver does not maintain arc consistency or use a
propagation queue).

@@

Solution-biased search allows for a much simpler parallel implementation. We
create a number of threads, and give each thread its own random seed, otherwise each thread
performs the same sequential search. Then, whenever a thread restarts, it uses a barrier to wait for
every other thread to also restart. At each restart each thread's new nogood can be gathered, and
then each thread resumes, with a larger set of nogood than in a sequential run. Finally, whenever
any single thread terminates, either due to having found a solution or proved unsatisfiability,
then every other thread may immediately terminate.

@@

\begin{itemize}
    \item Each thread will be run with the same restart schedule, and so will spend approximately
        the same amount of time between restarts. Because the only synchronisation between threads
        is at a restart, we expect threads to be busy doing search.  (This is in contrast
        to an alternative method for paralellising restarting search \cite{DBLP:conf/aaai/CireKS14},
        which packs together successive sequence values to produce a balanced workload.)
    \item Sequentially, on non-trivial instances the algorithm will restart often (many
        tens of thousands, for instances that reach the thousand second timeout).
    \item Because the search trees we explore are exponentially large, the randomness in the
        value-ordering heuristic is sufficient to ensure that most of the time, threads are
        exploring different parts of the search tree.
    \item The gathering of nogoods to describe the work done so far provides an alternative to
        requiring either a specific mechanism to allocate work, or expensive synchronisation between threads.
    \item If sometimes threads do happen to explore part of
        the same subproblems, this is not a problem: if the instance is satisfiable, either thread
        might find a solution first, and if the instance is unsatisfiable, we merely introduce some
        redundancy into the proof.\footnote{For solving a counting or enumeration problem,
        matters become slightly more complicated, but not devastatingly so.} The combination of
        rapid restarts and nogood recording is enough to ensure that this is only a small overhead.
\end{itemize}

@@

It is important to emphasise that there is no expectation of a linear speedup from this approach,
particularly for satisfiable instances. Due to the biased random nature of the value-ordering heuristic,
one of our extra threads may ``get lucky'' and find a solution much quicker than we would
sequentially, leading to a superlinear speedup. Conversely, our extra threads may contribute nothing to finding a
solution---or worse, due to new nogoods altering the choice made by the biased random branching heuristic,
we could even get an absolute slowdown. We can see both of these effects in the bottom left plot of
\cref{figure:parallel}: although we achieve roughly a linear speedup on unsatisfiable instances,
satisfiable instances show much greater variability. However, this approach does at least mirror our intuition of
allocating search effort in proportion to where the value-ordering heuristic believes it will be
most fruitful, and so we should not be too surprised that we see roughly a linear speedup on average
on harder instances.

@@

The authors would like to thank \"{O}zg\"{u}r Akg\"{u}n, Chris Jefferson, Christophe Lecoutre, Christine
Solnon, and Craig Reilly for their comments.
